---
title: "Gaussian GLMs in R"
format: html
---

# Data Generation 

From the introduction we created some fake/simulated data where we know there are differences 

We will try analysing this data first, then we will download some real-life data to also analyse.


```{r}
library(tidyverse)

Year <- seq(from=1950,to=2023,by=1)

Treatment <- c("Control","Treatment 1","Treatment 2")

Rep<- seq(from=1,to=10,by=1)

df1<-expand.grid(Year=Year,Treatment=Treatment,Rep=Rep)


Response<-rnorm(n=nrow(df1),mean = 15,sd=8)

df1$Response<-Response
```

This time we will replace the response column rather than making a new one

```{r}

df<-df1 %>% 
  mutate(Response=case_when(Treatment=="Control"~jitter(Response)*((Year-1930)/5)-50,
                                    Treatment=="Treatment 1"~jitter(Response)*((Year-1930)*(-2))+1000,
                                    Treatment=="Treatment 2"~jitter(Response)*((Year-1930)*1.5)-4))
```

# Modelling

Modelling wise we may have multiple different expectations from this data, are all treatments changing with time? in the same way?
The most complex model would be the interaction of Treatment and Year

Which would assume that the effect of Year is different depending on the Treatment level 
(decrease over time in one treatment, increase over time in another)

```{r}
glm1.1<-glm(Response~Treatment*Year,df, family=gaussian)
```

This is the same as using lm() and not specifying the family, as it is less code we will do this for gaussian models

```{r}
lm1.1<- lm(Response~Treatment*Year,df)
```

We can now check the residuals show no patterns

```{r}
plot(lm1.1)
```

This is annoying as we have to press enter in the console to see all the plots
We will install some packages from the easystats ecosystem of packages for this. There are other packages we could also use for this (including one of my own) but performance is good

```{r}
#| fig.width: 8
#| fig.height: 8

#install.packages("performance")
library(performance)

check_model(lm1.1)
```

That's better, we now have some interesting plots but our model doesnt look very good (Probably as we have simulated the data itself)

also we can look at the effect of Year and Treatment where there is no interaction
This equates to their being an overall effect of Year and then an effect of Treatment

```{r}
#| fig.width: 8
#| fig.height: 8

lm1.2<- lm(Response~Treatment+Year,df)

check_model(lm1.2)

lm1.3<- lm(Response~Year,df)

check_model(lm1.3)

lm1.4<- lm(Response~Treatment,df)

check_model(lm1.4)

lm1.5<- lm(Response~1,df)

check_model(lm1.5)
```

None of these model assessments are any good so we shouldn't draw any conclusions from any of these results but we shall try the same method with some real world data and then make some decisions.

Packages often have their own example datasets within them, or a package can be used to store just data without functions etc.

We will look at the palmer penguins dataset

# Real World Data

```{r}
#install.packages("palmerpenguins")

library(palmerpenguins)
data(penguins)
```

This becomes a 'promise' of a data set, we have to do something with it to get it properly, lets take a look inside

```{r}
glimpse(penguins)
```

This gives us two datasets in our global environment

As we are going to make simple models here we will clean the data to make our lives easier, this will mean removing NA values.
Removing NA's would normally be something I would not recommend, we (or someone) has worked extremely hard to create this data, 
so we shouldn't ever remove data, but NA replacement is complicated and not the subject for now.

Using summary() we can see which columns have NAs and which don't.

```{r}
summary(penguins)
```

Through going into the object and ordering one of the columns we find that the NA's (which order to last always) are the same rows
So we only have to remove two rows to remove the NAs in the biometrics columns, there are more NA's in the sex column
if we want to remove NA's there are many ways, to be selective we can filter our dataset 

Here we will make use of ! this means the opposite of the clause (not this)
We also use %in% which is used to tell filter there are more than one element
or NAs that we want to get rid of as NA is not classed like normal data

```{r}
penguins_someNAs<-penguins %>% 
  filter(!body_mass_g%in%NA)
```

If we now look at the number of rows of the datasets we can see only two rows were removed. (not all the NAs)

```{r}
nrow(penguins)
nrow(penguins_someNAs)

summary(penguins_someNAs)
```

Still 9 NAs in sex

```{r}
penguins_noNAs<-penguins_someNAs %>% 
  filter(!sex%in%NA)

summary(penguins_noNAs)
```

All sorted

# Modelling Categorical Factors

So now we will try prove the obvious
Does the flipper length of penguins change between species and between sexes
Whether we use an interaction or not depends on if our scientific thought 
believes the relationship of Species to flipper length is different between sexes (sexual dimorphism may not be consistent across species)


```{r}
#| fig.width: 8
#| fig.height: 8

lm2.1<-lm(flipper_length_mm~species*sex,data=penguins_noNAs)

check_model(lm2.1)
```

As we only have factors in our model we don't see a 'cloud' of points, but the line still is flat and horizontal so this is good
As the diagnostics are good we can look at the results

```{r}
summary(lm2.1)
```

Okay there are a lot of numbers here but what does it actually mean
first lets plot the raw data, boxplots are probably the best for categorical factors
We can re-use some of our code from the intro for appearance and colours

```{r}
#| fig.width: 8
#| fig.height: 8

ggplot(penguins_noNAs)+
  geom_boxplot(aes(x=species,y=flipper_length_mm,fill=sex))+
  scale_fill_manual(values=c("darkcyan","darkorange"))+
  labs(x="Sex",y="Response Variable (Flipper Length (mm))")+
  theme_classic()
```

Now we can also see what the model believes about our data 
This should be similar to our raw data but not identical
to do this we make simulated raw data with this same predictor variables in
we then use the model to predict the response variable based on those predictor variables

Therefore, we make a data set with just sex and species (be careful of spelling and capitalisation, R wants it identical)
the model then predicts the average Flipper length in mm based on those species and sexes. 
We can also tell the Predict function to predict error (Standard Error here)


```{r}
#| fig.width: 8
#| fig.height: 8

NewData<-expand_grid(sex=c("male","female"),
                     species=c("Adelie","Chinstrap","Gentoo"))

Pred<-predict(lm2.1,NewData,se.fit=TRUE)

NewData$response<-Pred$fit

NewData$se.fit<-Pred$se.fit


ggplot(NewData)+
  geom_point(aes(x=species,y=response,colour=sex),
             position=position_dodge(0.8))+
  geom_errorbar(aes(x=species,ymax=response+se.fit,
                    ymin=response-se.fit,colour=sex),
                width=0.1,
                position=position_dodge(0.8))+
  scale_colour_manual(values=c("darkcyan","darkorange"))+
  labs(x="Sex",y="Response Variable (Flipper Length (mm))")+
  theme_classic()
```


# Modelling Continuous Variables

Okay that is what we do with linear models of categorical factors
But what if we want to see the relationship between flipper_length_mm and bill_length_mm
But we know there are species differences and sexual differences in flipper length
As Males always tend to be larger lets just assess species differences in their flipper to bill relationship


Lets plot the raw data first 


```{r}
#| fig.width: 8
#| fig.height: 8

ggplot(penguins_noNAs)+
  geom_point(aes(x=bill_length_mm,y=flipper_length_mm,colour=species))+
  scale_colour_manual(values=c("darkcyan","darkorange","grey30"))+
  labs(x="Bill Length (mm)",y="Flipper Length (mm)")+
  theme_classic()
```

We can see from the raw data that we will expect to find some strong linear relationships

```{r}
#| fig.width: 8
#| fig.height: 8

lm3.1<-lm(flipper_length_mm~species*bill_length_mm,data=penguins_noNAs)

check_model(lm3.1)

summary(lm3.1)
```

As we hypothesised before modelling that there would be different bill to flipper relationships between species
The interaction model follows our scientific assumptions
Therefore, it would be incorrect to use lower complexity models (without the interactoin for example)

To predict again we want to create lines for each species
To do this we want to create fake bill length data over the same range for each species

Here we will use the seq() function that creates a sequence of values from your first number to your last number 
And you can chose the length of the vector it creates or the distance between each individual value


```{r}
NewData_<-expand_grid(bill_length_mm=seq(from=min(penguins_noNAs$bill_length_mm),
                                        to=max(penguins_noNAs$bill_length_mm),
                                        length.out=1000),
                     species=c("Adelie","Chinstrap","Gentoo"))
```

As the different species won't be across all of these bill length ranges 
we should also remove values outside of each species range

```{r}
Gentoo_Range<-penguins_noNAs %>% 
  filter(species=="Gentoo") %>% 
  summarise(min=min(bill_length_mm),
            max=max(bill_length_mm))

Adelie_Range<-penguins_noNAs %>% 
  filter(species=="Adelie") %>% 
  summarise(min=min(bill_length_mm),
            max=max(bill_length_mm))

Chinstrap_Range<-penguins_noNAs %>% 
  filter(species=="Chinstrap") %>% 
  summarise(min=min(bill_length_mm),
            max=max(bill_length_mm))


NewData_2<-NewData_ %>% 
  mutate(Range=case_when(species=="Gentoo" &
                              bill_length_mm>=Gentoo_Range$min &
                              bill_length_mm<=Gentoo_Range$max~"Good",
                         species=="Adelie" &
                           bill_length_mm>=Adelie_Range$min &
                           bill_length_mm<=Adelie_Range$max~"Good",
                         species=="Chinstrap" &
                           bill_length_mm>=Chinstrap_Range$min &
                           bill_length_mm<=Chinstrap_Range$max~"Good"
  )) %>% 
  filter(!Range%in%NA) %>% 
  select(-Range)
```

As we didnt create a case when for outside of the good range then they will be NAs and we can filter them out
Then we remove the created Range column

```{r}
Pred_2<-predict(lm3.1,NewData_2,se.fit=TRUE)

NewData_2$response<-Pred_2$fit

NewData_2$se.fit<-Pred_2$se.fit
```

So now we have many data points that can be used to draw the linear model outputs

```{r}
#| fig.width: 8
#| fig.height: 8

ggplot()+
  geom_ribbon(data=NewData_2,mapping=aes(x=bill_length_mm,ymax=response+se.fit,
                                                 ymin=response-se.fit,fill=species),
              alpha=0.4)+
  geom_line(data=NewData_2,mapping=aes(x=bill_length_mm,y=response,colour=species),
             alpha=0.4)+
  scale_color_manual(values=c("darkcyan","darkorange","grey30"))+
  scale_fill_manual(values=c("darkcyan","darkorange","grey30"))+
  labs(x="Bill Length (mm)",y="Response Variable (Flipper Length (mm))")+
  theme_classic()
```

This looks good but lets maybe add the raw data values onto the same figure as the model outputs

```{r}
#| fig.width: 8
#| fig.height: 8

ggplot()+
  geom_point(data=penguins_noNAs,mapping = aes(x=bill_length_mm,
                                               y=flipper_length_mm,
                                               colour=species),
             alpha=0.4,size=0.8)+
  geom_ribbon(data=NewData_2,mapping=aes(x=bill_length_mm,ymax=response+se.fit,
                                         ymin=response-se.fit,fill=species),
              alpha=0.4)+
  geom_line(data=NewData_2,mapping=aes(x=bill_length_mm,y=response,colour=species),
            alpha=0.4)+
  scale_color_manual(values=c("darkcyan","darkorange","grey30"))+
  scale_fill_manual(values=c("darkcyan","darkorange","grey30"))+
  labs(x="Bill Length (mm)",y="Response Variable (Flipper Length (mm))")+
  theme_classic()
```






